---
title: "Female Participation in the Workforce"
author: "Katie Braden (kbrad12@uw.edu), Josef Oscarsson (joscar@uw.edu), Jordan Kauffman (jbkauff@uw.edu)"
date: "11/14/22"
output: html_document
---

## ***Dynamic Paragraph:***
When looking at the seven values calculated above, it is important to realize what the 
significance of each of them have in helping to gain a better understanding of our data set and how it
can be applied within the scope of our project. The first value we calculated was in regard to the 
number of observations within the aggregated table set, which came out to 7,344. The next value was 
calculated using the 'unique' function to show the names of the various countries within the data set.
We also thought it was important to have a higher-level number just showing the amount of countries
within the data set. Due to this, we used the 'n_distinct' function and found that there are 272
countries. The fourth value we calculated was what country had the highest unemployment rate of 
women. This information was included in the 'summary_table_3' data set we aggregated and found that 
Algeria had the highest unemployment rate of women, in 1996. We decided to pull both 'Year' and 'Entity'
as it gives more information that leads to deeper insights. On the flip, we also calculated the 
country with the lowest unemployment rate of women using a similar function and found it returned
Cambodia and 2015. The sixth value was pertaining to the highest long run perspective on female
labor force participation rates. We found that the country with the highest long run perspective
was Iceland in 2016. Similar to the last, we also calculated the country with the lowest long
run perspective on female labor force participation rates and found that it was Turkey in 2005. All of
this data was compiled within the 'summary_info' list and helped us gain a better understanding of
our data sets and some of the clear patterns that can be found.

## ***Aggregate Table Script:***
```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyverse)
library(knitr)
unemployment_rate_women <- read.csv("C:/Users/katie/Documents/info201/assignments/project-group-3-section-ad/data/unemployment-rate-women.csv")

names(unemployment_rate_women) [4] = "Unemployment_rate_women"

unemployment_rate_women$Unemployment_rate_women <- round(unemployment_rate_women$Unemployment_rate_women, digit=2)

long_run_perspective <- read.csv("C:/Users/katie/Documents/info201/assignments/project-group-3-section-ad/data/Long run perspective on female labor force participation rates.csv")

names(long_run_perspective) [4] = "long_run_perspective_female_labor_force_participation_rate"


female_labor_force_by_age <- read.csv("C:/Users/katie/Documents/info201/assignments/project-group-3-section-ad/data/female-labor-force-by-age.csv")

names(female_labor_force_by_age) [4] = "65+_females_in_labor_force"
names(female_labor_force_by_age) [5] = "55-64_females_in_labor_force"
names(female_labor_force_by_age) [6] = "45-54_females_in_labor_force"
names(female_labor_force_by_age) [7] = "35-44_females_in_labor_force"
names(female_labor_force_by_age) [8] = "25-34_females_in_labor_force"
names(female_labor_force_by_age) [9] = "15-24_females_in_labor_force"

ratio_of_female_to_labor_force_participation_rates <- read.csv("C:/Users/katie/Documents/info201/assignments/project-group-3-section-ad/data/ratio-of-female-to-male-labor-force-participation-rates.csv")

names(ratio_of_female_to_labor_force_participation_rates) [4] = "ratio_of_female_to_male_labor_force_participation_rate"

ratio_of_female_to_labor_force_participation_rates$ratio_of_female_to_male_labor_force_participation_rate <- round(ratio_of_female_to_labor_force_participation_rates$ratio_of_female_to_male_labor_force_participation_rate, digit=2)

summary_table_1 <- left_join(unemployment_rate_women, long_run_perspective, group_by= c("Year", "Entity", "Code"), all.x=TRUE)
  
summary_table_2 <- left_join(summary_table_1, female_labor_force_by_age, group_by = c("Year", "Entity", "Code"), all.x=TRUE)

summary_table_3 <- left_join(summary_table_2, ratio_of_female_to_labor_force_participation_rates, group_by=c("Year", "Entity", "Code"), all.x=TRUE)

names(summary_table_3) [1] = "Country"
```

### *Table is Abbreviated to First 10 Values*
```{r, warning=FALSE, message=FALSE}
head(summary_table_3, 10)
```

### *Aggregate Table Explanation:*
This table serves as a summary of the information and data that we have chosen to highlight in our report and include in our visualizations. The common variables in this table are 'Country', 'Code', and "Year'. Besides these entities, the important columns that this table includes are 'Unemployment_rate_women', 'long_run_perspective_female_labor_force_participation_rate', females in labor force by age (ie. '65+_females_in_labor_force), and 'ratio_of_female_to_male_labor_force_participation_rate'. We chose to include these pieces of data because we thought they best reflected on both the historical and current state of female participation in the workforce and how that compares to males. For brevity in this report, we chose to only show the first 10 values in the markdown file but refer to the summary information for the actual length of rows in the table (272). 


## ***Summary Information Script:***
```{r, warning=FALSE, message=FALSE}

library(dplyr)
library(tidyverse)
#List Containing Different Values from Data Set
summary_info <- list()

# 1 - Number of Observations 
summary_info$number_observations <- nrow(summary_table_3)
summary_info$number_observations

# 2 & 3 - Number of Unique Countries/Entities in Dataset
summary_info$list_of_countries <- unique(summary_table_3$Country)

summary_info$number_of_countries <- length(unique(summary_table_3$Country))
summary_info$number_of_countries

#4 & 5 - Highest/Lowest Unemployment Rate of Women
summary_info$highest_unemployment_rate_of_women <- summary_table_3 %>% 
  filter(Unemployment_rate_women == max(Unemployment_rate_women, na.rm = TRUE)) %>% 
  pull(Year,Country)
summary_info$highest_unemployment_rate_of_women

summary_info$lowest_unemployment_rate_of_women <- summary_table_3 %>% 
  filter(Unemployment_rate_women == min(Unemployment_rate_women, na.rm = TRUE)) %>% 
  pull(Year,Country)
summary_info$lowest_unemployment_rate_of_women

# 6 & 7 -Highest/Lowest Long Run Perspective on Female Labor Force Participation Rates
summary_info$highest_long_run_perspective <- summary_table_3 %>% 
  filter(long_run_perspective_female_labor_force_participation_rate == 
           max(long_run_perspective_female_labor_force_participation_rate,
               na.rm = TRUE)) %>% 
  pull(Year, Country)
summary_info$highest_long_run_perspective

summary_info$lowest_long_run_perspective <- summary_table_3 %>% 
  filter(long_run_perspective_female_labor_force_participation_rate == 
           min(long_run_perspective_female_labor_force_participation_rate,
               na.rm = TRUE)) %>% 
  pull(Year, Country)
summary_info$lowest_long_run_perspective

#Summary Information List
View(summary_info)
```

## ***Chart 1:***
```{r, fig.width=12, fig.height=6,warning=FALSE, message=FALSE}
chart1 <- ggplot() +
  geom_line(aes(y=long_run_perspective_female_labor_force_participation_rate, x=Year,group=Entity, color=Entity), data = long_run_perspective) +
  ggtitle("Long-run Perspective on Female Labor Force Participation Rates", subtitle= "Proportion of the Female Population Ages 15 and Over that is Economically Active") +
  xlab("Year") +
  ylab("Female Labor Force Participation Rates (%)")

chart1
```

### *Chart 1 Explanation:*
We chose to incorporate this graph into our research as we believe it accurately
depicts the long-run perspective on female labor force participation rates
over the course of the past 100 years. We chose to include many countries
in this visualizaiton as we believe it encompasses broader trends across
the world. One observation from the graph is that a majority of the data
was compiled beginning in the 1960-70s, with very few countries having data
that dates back prior to the 1920s. Another observation is that nearly every
country has experienced rapid growth in the percentage of the female population,
15+, that is economically active compared to the growth rates in the earlier
periods, such as 1920-1960.

## ***Chart 2:***
```{r, fig.width=12, fig.height=6, warning=FALSE, message=FALSE}
library(maps)
mapdata <- map_data("world")

View(mapdata)

View(unemployment_rate_women)

names(unemployment_rate_women) [1] = "region"

unemployment_rate_women <- subset(unemployment_rate_women, region!= "Americas: High income" & region!= "Americas: Low income" & region!= "Americas: Lower-middle income" & region!= "Americas: Upper-middle income" & region!= "Africa: Low income" & region!= "Africa: Lower-middle income"& region!= "Africa: Upper-middle income" & region!= "Americas") 

View(unemployment_rate_women)

unemployment_rate_women_chart <- read.csv("C:/Users/katie/Documents/info201/assignments/project-group-3-section-ad/data/unemployment-rate-women-chart.csv") %>% 
  filter(Year==2017)

View(unemployment_rate_women_chart)

mapdata <- left_join(mapdata,unemployment_rate_women_chart, group_by= "region")

mapdata1 <- mapdata %>% 
  filter(!is.na(mapdata$Unemployment_rate_women))

map1 <- ggplot(mapdata1, aes( x = long, y= lat, group=group)) +
  geom_polygon(aes(fill = Unemployment_rate_women), color = "black")

map2 <- map1 + scale_fill_gradient(name = "Female Unemployment Rate (%)", low = "pink", high = "red", na.value = "grey50")

chart2 <- map1 + scale_fill_gradient(name = "Female Unemployment Rate (%)", low = "pink", high = "red", na.value = "grey50")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank (),
        axis.title.y = element_blank(),
        axis.title.x = element_blank (),
        rect = element_blank()) +
  ggtitle("Unemployment Rate, Women, 2017", subtitle= "The unemployment rate expresses the number of unemployed as a percent of the labor
force. Countries in white have no unemployment data recorded.")
chart2
```

### *Chart 2 Explanation:*
We decided to include a chart of the unemployment rate with the most recent data (2017) to show the current (or most recent) 
state of female unemployment. The other data sets displays trends over time so this data set summarizes that state that those 
long-term trends have lead to. Unemployment data is a major indicator of job certainty and equality so we chose
this feature as a key one to highlight. The observations include unemployment data from 2017 from most countries in
the world. The countries that were not documented in the data set are displayed in white on the graph. This map is in the form
of a heat map where the more red the countries are, the higher unemployment rate they have, and the more pink the 
countries are, the lower unemployment rate they have.

## ***Chart 3:***
```{r, fig.width=12, fig.height=6, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)

chart_3_data_set_complete <- read.csv("C:/Users/katie/Documents/info201/assignments/project-group-3-section-ad/data/female-labor-force-by-age-chart-3-v2.csv")
View(chart_3_data_set_complete)

chart3 <- ggplot(chart_3_data_set_complete, aes(x = Year, y = Number, fill = Age)) +
  geom_area(colour = "black", size = .2, alpha = .4) +
  scale_fill_brewer(palette = "Reds") +
  geom_line(position = "stack", size = .3) +
  labs(title = "Female Labor Force by Age",
     x = "Year",
     y = "Number of Females (In Millions)") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank()) 
chart3
```

### *Chart 3 Explanation:*
We chose to incorporate this graph into our research as we believe it does 
a great job of showing the changes in the number of females in the labor
force, over the course of 1990 through 2030 (projected). Through the stacked
area chart, it very clearly shows the changes in the labor force numbers
by age group, with the legend/color scale directly to the right of the chart.
One immediate observation from the chart is that groups, particularly ages
15-24, 25-34, and 35-44 have experienced the highest growth rates over the 
period.

